Safe Memory Regions for Big Data Processing

### 2016-06-10 paper/type-inference.tex
type inference

\begin{itemize}
\item Add destructive object-field-update \emph{x.f := y} to the language.
\item Add if-then-else to the language.
\item Omit constructors from classes (and use an implicit constructor).
\item Note that the claim that all region variables will be unified with some concrete
region fails in the case of a method that recursively calls itself (and, hence, does not
terminate). In this case, the region variables in the return type are not constrained.
Similarly for non-terminating mutual recursion. These are uninteresting (in essence, the
return type can be taken to be unit instead).
\item Discuss the safety of virtual method invocation, which provides a limited form
of existentially quantifying regions and instantiating (unpacking) them because a
derived class may have more region parameters than a base class.
\item Discuss the subtyping rule used to check an overriding method (in the static
semantics).
\item Mention that constraints are in the theory of partial orders?
\end{itemize}

@@

\textbf{TODO: The following not yet integrated into above.}

@@

This iterative process is a standard fixed point computation process.
It is possible to encode this computation using a set of Datalog rules, and compute the
solution using any Datalog engine.
Alternatively, note that the above computation, in the absence of context constraints
$\phictxt$ on the left-hand-side of the entailment, is a special instance of context-free
reachability in graphs. Algorithms for context-free reachability can be adapted to
incorporate the above treatment of context constraints $\phictxt$.

@@

\subsubsection{Non-Recursive Constraints}
\label{sec:csolve}

We first describe how we solve a non-recursive CSP $\csp$ where
$c\lbrack \varphi,\rhobar \rbrack$ is of the form $\isvalid{\Phicx
\conj \varphi}{\Phics}$.

Our approach is based on the observation that $\FB$ does not admit
null values or uninitialized variables, thus forcing every region
variable to be unified with some concrete region.
%
The constraint formula $\Phics$ captures all such unification constraints on $\rhobar$.
Since the set of all concrete region identifiers is the
unification domain ($\rhoenv$) of CSP, this means for every $\rho_i$
with a well-formedness constraint as $\rho_i \in \rhoenv_i$, there
exists a $\rgn \in \rhoenv$ such that and $\isvalid{\Phicx \conj
\Phics}{\rho = \rgn}$. However, $\rgn$ may not belong to $\rhoenv_i$,
in which case the well-formedness constraint on $\rho_i$ is not
satisfied, and constraint solving must fail.  Therefore, there exists
a unique assignment $\substFn$ such that $c\lbrack
\phi_{sol},\substFn(\rhobar)$ is satisfied, regardless of
$\phi_{sol}$.

To obtain $\phisol$, we make use of another observation. Let $c\lbrack
\varphi, \substFn(\rhobar) \rbrack$ be the following constraint:
\begin{center}
\(
  \isvalid{ \Phicx \conj \varphi}
        {\Phics'}
\)
\end{center}
Consider a maximally weak formula $\phi$ such that
$\isvalid{\Phicx}{\phi \Leftrightarrow \Phics'}$. 
% Since $\phi$ is
% maximally weak, there does not exist a $\phi'$ satisfying all the
% below three conditions:
% \begin{itemize}
% \item $\isvalid{\Phicx}{\phi' \Leftrightarrow \Phics'}$.
% \item $\isvalid{\cdot} {\phi \Rightarrow \phi'}$.
% \item $\isnotvalid{\cdot}{\phi' \Rightarrow \phi}$.
% \end{itemize}
Clearly, $\tywf{\rhoenv}{\phi}$. However, since $\rhoenv_{\varphi}
\subseteq \rhoenv$, we have two cases:
\begin{itemize}

\item Case $\tywf{\rhoenv_{\varphi}}{\phi}$: This means that $\phi$ is
a solution to $\varphi$. 
% Furthermore, it is the weakest solution, because it is equivalent to
% $\Phics$ under $\Phicx \conj \bigwedge_j(\rho_j =
% \substFn(\rho_j))$, and any other solution has to stronger than
% $\Phics$ under the same context.

\item Case $\tynwf{\rhoenv_{\varphi}}{\phi}$: In this case, $\phi$
contains at least one equality or outlives constraint on two region
identifiers, $\rgn_i$ and $\rgn_j$, where (a). $\rgn_i,\rgn_j \in
\rhoenv - \rhoenv_{\varphi}$, or (b). $\rgn_i \in \rhoenv -
\rhoenv_{\varphi}$ and $\rgn_j \in \rhoenv_{\varphi}$, such that the
constraint is not implied by $\Phicx$ (if it is implied, then $\phi$
is not maximally weak). Let us denote such constraint on $\rgn_i$ and
$\rgn_j$ as $\phi_{ij}$. Now, let us consider a solution $\phisol$ to
$\varphi$, which means that $\isvalid{\Phicx \conj \phisol}{\Phics'}$.
Since $\isvalid{\Phicx}{\phi \Leftrightarrow \Phics'}$, we have
$\isvalid {\Phicx \conj \phisol}{\phi}$. Since
$\isvalid{\phi}{\phi_{ij}}$, we have $\isvalid {\Phicx \conj
\phisol}{\phi_{ij}}$, although $\isnotvalid{\Phicx}{\phi_{ij}}$. But
this is impossible. To see why, recall that all the constraints on
identifiers in $\rhoenv_{\varphi}$ occuring in $\Phicx$ are subsumed
by $(\rhoenv - \rhoenv_{\varphi}) \outlives \rhoenv_{\varphi}$.
Therefore, it is impossible to derive $\phi_{ij}$ from $\Phicx$ by
only adding constraints on $\rgn \in \rhoenv_{\varphi}$. Hence, such a
solution $\phisol$ cannot exist.
\end{itemize}

@@

The above discussion hints at an algorithm to compute a solution to
$\varphi$: find a maximally weak $\phisol$ such that $\isvalid{\Phicx}
{\phisol \Leftrightarrow \Phics'}$. If $\tywf{\rhoenv_{\varphi}}
{\phisol}$, then $\phisol$ is the solution. Otherwise, there is no
solution to $\varphi$. We now describe a graph-based algorithm to
compute maximally weak $\phisol$.

\paragraph{Definitions} Given a constraint formula $\phi$, we define
its graph encoding $G(\phi)=(V(\phi),E(\phi))$ as a digraph whose
vertices ($V(\phi)$) are free region variables and identifiers in
$\phi$, and whose edges ($E(\phi)$) denote outlives constraints
in\footnote{We alternate between viewing a constraint formula ($\phi$)
as a set of constraints and a conjunction of constraints in this
description} $\phi$.  That is, if $\phi$ contains a constraint $\rho_1
\outlives \rho_2$, then $\rho_1,\rho_2 \in V(\phi)$ and
$(\rho_1,\rho_2) \in E(\phi)$. Equality constraints are treated as a
conjunction of symmetric outlives constraints for the purpose of graph
encoding. Conversely, given a digraph $G$, we define its constraint
encoding $\Phi(G)$ in a straightforward manner. We say that a graph
$G_1=(V_1,E_1)$ is as connected as graph $G_2=(V_2,E_2)$ if $V_2
\subseteq V_1$, and for every $\rho_1, \rho_2 \in V_2$, if there
exists a path between $\rho_1$ and $\rho_2$ in $G_2$, then there must
exist a path between same pair of vertices in $G_1$.

A maximally weak $\phisol$ that satisfies $\isvalid{\phictxt}{\phisol
\Leftrightarrow \phicstr}$ is a constraint encoding of the smallest
graph $G$ (i.e., $\Phi(G)$) such that $G(\phictxt) \cup G$ is as
connected as $G(\phicstr)$. The problem of computing such a $G$ is
equivalent to the problem finding of finding minimum number of edges
to add to $G(\phictxt)$ such that it is as connected as $G(\phicstr)$.
Algorithms to solve the later problem are known to
exist~\cite{siam92}.

Solving the constraints $c_{12}$ and $c_{13}$ by reducing them to
graph augmentation problems, as described above, results in the
following solutions for $\varphi_0$ and $\varphi_1$ (symmetric
outlives constraints are replaced with equalities):
\begin{center}
\(
  \varphi_1 \Leftrightarrow \rho_0 \outlives \rhoalloc_0 \conj \rho_1
     \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0 \conj 
     \rho_2 = \rho_0 \conj \rho_3 = \rho_1
\)\\
\(
  \varphi_2 \Leftrightarrow \rho_5 = \rho_0
\)
\end{center}

\subsubsection{Solving Recursive Constraints}

\textbf{TODO: FP iteration should use $\Phicx$}

Substituting the above solutions for $\varphi_0$ and $\varphi_1$ in
$c_{14}$ gives us a recursive constraint of the following form:
\begin{center}
\(
  \isvalid{\phictxt \conj \varphi}{\phicstr \conj F(\varphi)}
\)
\end{center}
Where $\phictxt$ and $\phicstr$ are concrete constraint formulas, and
$F$ is a substitution function, not necessarily idempotent. We now
extend constraint solving to recursive CSP $\csp$, where $c\lbrack
\varphi,\rhobar \rbrack$ assumes the form $\isvalid{\Phicx \conj
\varphi}{\Phics \conj \bigwedge_i F_i(\varphi)}$. For the sake of
brevity, we define $G(\varphi) = \Phics \conj \bigwedge_i
F_i(\varphi)$, and use $F$ in $c\lbrack \varphi,\rhobar \rbrack$: 
\begin{center}
\(
  \isvalid{\Phicx \conj \varphi}{G(\varphi)}
\)
\end{center}
To solve the above recursive constraint, we start with the observation
that the set of all constaints ($\phi$) over $\rhoenv \cup
\{\rhobar\}$ is a lattice, where $\phi_1 \le \phi_2 \triangleq
\isvalid{\cdot}{\phi_1 \Rightarrow \phi_2}$. Note that $G$ is a
monotone over the lattice:
\begin{center}
  $\forall \phi_1,\phi_2.\; \isvalid{\cdot}{(\phi_1 \Rightarrow
  \phi_2) \Rightarrow (G(\phi_1) \Rightarrow G(\phi_2))}$
\end{center}
From Knaster-Tarski's theorem, we know that $G$ has a greatest fixed
point ($\phi_f$), with following properties:
\begin{itemize}
\item \textbf{Property 1} $\phi_f = G(\phi_f)$, which means
$\isvalid{\cdot}{\phi_f \Leftrightarrow G(\phi_f)}$
\item \textbf{Property 2} $\forall \phi_f'$ such that $\phi_f' =
G(\phi_f')$, we have $\phi_f' \le \phi_f$, which means $\isvalid
{\cdot} {\phi_f' \Rightarrow \phi_f}$.
\end{itemize}
We therefore compute the greatest fixed point ($\phi_f$) of $G$, and
convert the recursive constraint to the following non-recursive
constraint:
\begin{center}
\(
  \isvalid{\Phicx \conj \varphi}{\phi_f}
\)
\end{center}
The technique described in \S~\ref{sec:csolve} now suffices to solve
the above constraint.

Let $H(\varphi_2)$ denote the consequent of the recursive constraint
$c_{14}$. Its greatest fixed point ($\phi_f$) is shown below:
\begin{center}
\(
  \rho_7 \outlives \rho_6 \conj \rho_8 \outlives \rho_6 \conj \rho_7 =
  \rho_9 \conj \rho_8 = \rho_9 \conj \rgn_0 \outlives \rgn_1 \conj
  \rgn_0 = \rgn_0
\)
\end{center}
Computing a maximally weak $\phisol$ such that $\isvalid{\varphi_1
\conj \rgn_0 \outlives \rgn_1}{\phisol \Leftrightarrow \phi_f}$
results in the following solution for $\varphi_2$ that meets its
well-formedness requirement ($\tywf{\{\rhoalloc_0,\rho_{}0-4,
\rhoalloc_2, \rho_{6-9}\}}{\varphi_2} $):
\begin{center}
\(
  \rho_7 \outlives \rho_6 \conj \rho_8 \outlives \rho_6 \conj \rho_7 =
  \rho_9 \conj \rho_8 = \rho_9
\)
\end{center}

% Note that $F$ is a non-idempotent substitution function with following
% properties:
% \begin{itemize}
% \item $F$ is a monotone: 
% \begin{center}
%   $\forall \phi_1,\phi_2.\; \isvalid{\cdot}{(\phi_1 \Rightarrow
%   \phi_2) \Rightarrow (F(\phi_1) \Rightarrow F(\phi_2))}$
% \end{center}

% \item $F$ distributes over conjunction (mind the syntactic equality):
% \begin{center}
%   $\forall \phi_1,\phi_2.\; F(\phi_1 \conj \phi_2) = F(\phi_1) \conj
%   F(\phi_2)$
% \end{center}
% \end{itemize}

% Define $G(\varphi) = \Phics \conj F(\varphi)$. Note that $G$ is also a
% monotone, hence $G$ has greatest fixed point (GFP) in the lattice. Let
% $\phi_f$ be the GFP of $G$. Following properties ensue:
% \begin{itemize}
% \item \textbf{Property 1} $\phi_f = G(\phi_f)$, which means
% $\isvalid{\cdot}{\phi_f \Leftrightarrow G(\phi_f)}$
% \item \textbf{Property 2} $\forall \phi_f'$ such that $\phi_f' =
% G(\phi_f')$, we have $\phi_f' \le \phi_f$, which means $\isvalid
% {\cdot} {\phi_f' \Rightarrow \phi_f}$.
% \end{itemize}

% \begin{theorem}
% $\phi_f$ is a solution.
% \end{theorem}
% \begin{proof}
% $\phi_f$ is GFP of $G$. Therefore, $\isvalid{\cdot}{\phi_f
% \Leftrightarrow G(\phi_f)}$. By strengthening the context, we get
% $\isvalid{\Phicx}{\phi_f \Leftrightarrow G(\phi_f)}$. It follows that
% $\isvalid{\Phicx \conj \phi_f}{G(\phi_f)}$. Hence, $\phi_f$ is a
% solution for $\varphi$.
% \end{proof}

% \begin{theorem}
% If $\phisol$ is a solution of $\varphi$, then $\isvalid{\Phicx \conj
% \phisol}{\phi_f}$
% \end{theorem}
% \begin{proof}
%   Since $\phisol$ is a solution, $\isvalid{\Phicx}{\phisol \Rightarrow
%   G(\phisol)}$. Therefore, $\isvalid{}{\Phicx \conj \phisol \Rightarrow
%   \Phicx \conj G(\phisol)}$. Therefore, $\isvalid{}{G(\Phicx) \conj
%   G(\phisol) \Rightarrow G(\Phicx) \conj G^2(\phisol)}$

% \end{proof}

>>>>

@@

@@

This iterative process is a standard fixed point computation process.
It is possible to encode this computation using a set of Datalog rules,
and compute the solution using any Datalog engine.
Alternatively, algorithms for context-free reachability can be adapted
to solve the constraints, as explained below.

Given a set $S$ of outlives constraints, we define the directed
graph $G(S)=(V(S),E(S))$ as follows.
Every distinct region identifier $\rho$ in $S$ is represented by a vertex,
which we will also refer to as $\rho$.
Every outlives constraint $\rho_1 \outlives \rho_2$ is represented by
an edge from $\rho_1$ to $\rho_2$.
Recall that an equality constraint $\rho_1 = \rho_2$ is shorthand
for the pair of constraints $\rho_1 \outlives \rho_2$ and
$\rho_2 \outlives \rho_1$: thus, an equality constraint is represented
by a pair of edges.
%
It is easy to see that $S \models \rho_1 \outlives \rho_2$ iff
there exists a path from $\rho_1$ to $\rho_2$ in $G(S)$.
%
Thus, simple graph reachability (transitive closure) algorithms can be
used to identify all the logical consequences of a set of outlives
constraints.

Given a set of context-free validity constraints, we can generalize the
above approach and reduce the problem to one of context-free reachability
in graph~\cite{Reps:Reachability} (as usual for context-sensitive interprocedural analysis).
%
Algorithms for context-free reachability can also be adapted to incorporate
the above treatment of context constraints $\phictxt$.
%
Further details omitted.

@@

\begin{itemize}
\item Add destructive object-field-update \emph{x.f := y} to the language.
\item Add if-then-else to the language.
\item Omit constructors from classes (and use an implicit constructor).
\item Discuss the safety of virtual method invocation, which provides a limited form
of existentially quantifying regions and instantiating (unpacking) them because a
derived class may have more region parameters than a base class.
\item Discuss the subtyping rule used to check an overriding method (in the static
semantics).
\item Mention that constraints are in the theory of partial orders?
\item Update discussion of examples
\end{itemize}

\textbf{TODO: Miscellaneous points to be integrated.}

To see why, recall that all the constraints on
identifiers in $\rhoenv_{\varphi}$ occuring in $\Phicx$ are subsumed
by $(\rhoenv - \rhoenv_{\varphi}) \outlives \rhoenv_{\varphi}$.
Therefore, it is impossible to derive $\phi_{ij}$ from $\Phicx$ by
only adding constraints on $\rgn \in \rhoenv_{\varphi}$. Hence, such a
solution $\phisol$ cannot exist.

@@

\subsection{Example}
